---
title: "MappingFutureOceanConditions"
author: "Max Kuempel"
date: "2025-06-01"
output: html_document
---
---
title: "Conditon Maps"
output: html_notebook
---

This notebook is used for mapping the future conditons of the Bering Sea. Bio-ORACLE layers are used (to match models). Layers are downloaded in the same way as in the model

For how to install bioracler, please consult the pollock modeling notebook under Working_Model. 

```{r Setup Packages}
library(biooracler)
library(terra)
library(ggplot2)
library(tidyterra) #Integrate terra plots into ggplot2
library(gridExtra)
```

Same contraints are used as in modeling for consistency
```{r Setup Constrains}
time = c('2010-01-01T00:00:00Z', '2010-01-01T00:00:00Z') #2010 layer date
latitude = c(52, 64) #latitude bounds of the 2010 Trawl data
longitude = c(-157, -179.975) #longitude bounds of the 2010 Trawl data
constraints2010 = list(time, latitude, longitude) #combines factors defined above
names(constraints2010) = c("time", "latitude", "longitude") #assigns names to each constraint so biooracler can read them


time2040 = c('2040-01-01T00:00:00Z', '2040-01-01T00:00:00Z')
constraints2040 = list(time2040, latitude, longitude)
names(constraints2040) = c("time", "latitude", "longitude")
```

```{r Temperature}
BenthicTempBaselineID <- "thetao_baseline_2000_2019_depthmax"
TempAvg2010  <- download_layers(BenthicTempBaselineID, 
                                variables = "thetao_mean", 
                                constraints = constraints2010)

BenthicTemp370ID <- "thetao_ssp370_2020_2100_depthmax"
TempAvg2040ssp370  <- download_layers(BenthicTemp370ID, 
                                variables = "thetao_mean", 
                                constraints = constraints2040)

BenthicTemp585ID <- "thetao_ssp585_2020_2100_depthmax"
TempAvg2040ssp585  <- download_layers(BenthicTemp585ID, 
                                variables = "thetao_mean", 
                                constraints = constraints2040)


```

```{r SeaIce}
IceThickBaselineID <- "sithick_baseline_2000_2020_depthsurf"
IceThickAvg2010  <- download_layers(IceThickBaselineID, 
                                variables = "sithick_mean", 
                                constraints = constraints2010)

IceThick370ID <- "sithick_ssp370_2020_2100_depthsurf"
IceThickAvg2040ssp370  <- download_layers(IceThick370ID, 
                                variables = "sithick_mean", 
                                constraints = constraints2040)

IceThick585ID <- "sithick_ssp585_2020_2100_depthsurf"
IceThickAvg2040ssp585  <- download_layers(IceThick585ID, 
                                variables = "sithick_mean", 
                                constraints = constraints2040)

```


```{r Plotting}
#Text size
 textscale <- theme(
    legend.title = element_text(size = 10),  # Legend label size
    legend.text = element_text(size = 10)    # Tick label size
  )

### Temperature ###
scale <- c(-1, 9)
colorscheme <-  scale_fill_gradient2(
    low = "darkblue",     # Low values
    mid = "lightblue",    # Midpoint
    high = "red",     # High values
    midpoint = 2,     # Center value
    name = "Temp (C)",
   limits = scale)

T2010 <- ggplot()+
  geom_spatraster(data = TempAvg2010)+
  colorscheme+
  labs(title = "2010")+
  theme_bw() + textscale


T2040_3 <- ggplot()+
  geom_spatraster(data = TempAvg2040ssp370)+
  colorscheme+
  labs(title = "2040 (SSP3-7.0)")+
  theme_bw() + textscale

T2040_5 <- ggplot()+
  geom_spatraster(data = TempAvg2040ssp585)+
  colorscheme+
  labs(title = "2040 (SSP5-8.5)")+
  theme_bw() + textscale

### Sea Ice Thickness ###
scale <- c(0, 0.5)
colorscheme <-  scale_fill_gradient2(
    low = "darkblue",     # Low values
    mid = "blue",    # Midpoint
    high = "white",     # High values
    midpoint = 0.1,     # Center value
    name = "Thickness (m)",
   limits = scale)

SI2010 <- ggplot()+
  geom_spatraster(data = IceThickAvg2010)+
  colorscheme+
  labs(title = "2010")+
  theme_bw() + textscale


SI2040_3 <- ggplot()+
  geom_spatraster(data = IceThickAvg2040ssp370)+
  colorscheme+
  labs(title = "2040 (SSP3-7.0)")+
  theme_bw() + textscale

SI2040_5 <- ggplot()+
  geom_spatraster(data = IceThickAvg2040ssp585)+
  colorscheme+
  labs(title = "2040 (SSP5-8.5)")+
  theme_bw() + textscale

### Plotting ###

grid.arrange(T2010, T2040_3,T2040_5,SI2010,SI2040_3,SI2040_5,
             nrow = 2)

```


